---
title: "ABCD NDA 2.0.1 activities questionnaire"
output: github_document
---

Analyze ABCD Activities questionnaire from DEAP 2.0.1 Rds file (doi:10.15154/1504431). No longer need to call this after running merge_data_music, core_demographics_music and categorical_extension_music
However, to simplify our life, may want to run subset_DEAP_Rds in order to create a smaller dataset, omitting columns of no (present) interest

Preamble
# Preamble
```{r}
rm(list=ls())

#library(dtplyr)
library(dplyr)
library(tidyr)
library(data.table)
library(gsubfn)
library(rlang)
library(corrplot)
library(tibble)
```

Load Data from 01_preprocessActivities_2.0.1.R, if needed
```{r}
# have to define the timestamp of data we wish to use
ts <- "2020-07-16_11-27-24"

# I/O
host = Sys.info()["nodename"]
if ( pmatch("amusing", host, nomatch=FALSE) ) {
  dataDir <- "C:/Users/jiversen/Documents/RDS/"
} else if ( pmatch("taiko", host, nomatch=FALSE) ) {
  dataDir <- "/Users/jri/Documents/ Research/Projects/simphony/ABCD/Data/"
}


resultDir = file.path(dataDir,"results",ts)

if (!"actData" %in% ls()) {
  activitiesFile      <- file.path(resultDir,"nda2.0.1_activities.Rds")
  activitiesFileLong  <- file.path(resultDir, "nda2.0.1_activities_long.Rds")
  activitiesFileMeta  <- file.path(resultDir, "nda2.0.1_activities_meta.RData")
  activitiesFileMusic <- file.path(resultDir, "nda2.0.1_activities_music.csv")

  actData <- readRDS(activitiesFile)
  actDataL <- readRDS(activitiesFileLong)
  load(activitiesFileMeta)
  musicData <- fread(activitiesFileMusic)
}

```


Plot basic facts about each activity

```{r}

activities <- unlist(strapplyc(names(actData),"(.*)_school$"))
measurements <- c("school",	"outside",	"private",	"self",	"nyr",	"nmonth",	"perwk",	"tspent",	"p12", "any", "ageStarted", "hrperwk", "hrlifetime")

for (act in activities) {
  
  actVars <- paste0(act,"_",measurements)
  adata <- actData[,actVars]
  
  par(mfrow = c(2, 2))
  par(cex = 0.6)
  par(mar = c(3, 3, 1, 1), oma = c(1, 1, 1, 1))
  hist(adata[[paste0(act,'_ageStarted')]], main=paste0(act,' ageStarted'),freq=FALSE)
  hist(adata[[paste0(act,"_hrperwk")]], main=paste0(act,' Hrs/wk'),xlim=c(0, 8),freq=FALSE)
  
  #get percentage participating in each context
  n=sum(adata[[paste0(act,'_any')]],na.rm=TRUE)
  percContext = apply(adata[,1:4],2,sum,na.rm=TRUE) / n
  barplot(percContext,names.arg=c("school","outside","private","self"),main="Contexts")
  
  #correlation of contexts
  mat = data.matrix(adata[,1:4])
  mat[is.na(mat)]=0
  cc = cor(mat)
  corrplot(cc,diag=FALSE)
  }
```

Plot participation in activities
# simple distribution of participants endorsing any type of a given activity and in past 12 months
```{r}
activities <- activities[1:29]
percAct <- 100*sapply(actData[paste0(activities,"_any")],sum,na.rm=TRUE) / nrow(actData)
percp12 <- 100*sapply(actData[paste0(activities,"_p12")],sum,na.rm=TRUE) / nrow(actData)

percSport <- 100*sum(actData$sports_any>0,na.rm=TRUE) / nrow(actData)
percSport_p12 <- 100*sum(actData$sports_p12>0,na.rm=TRUE) / nrow(actData)
percArts <- 100*sum(actData$arts_any,na.rm=TRUE) / nrow(actData)
percArts_p12 <- 100*sum(actData$arts_p12,na.rm=TRUE) / nrow(actData)
#percMusicNoSport <- 100*sum(actData$musicNoSport,na.rm=TRUE) / nrow(actData)
#percMusicNoSport_p12 <- 100*sum(actData$musicNoSport_p12,na.rm=TRUE) / nrow(actData)
aggPerc <- as.data.frame(cbind(percArts, percArts_p12, percSport, percSport_p12))

print("=== Activity endorsement: p12")
sort(percp12,decreasing = TRUE)
par(mar=c(11,4,4,4));barplot(sort(percp12,decreasing = TRUE),las=2)

print("=== Activity endorsement: Lifetime")
sort(percAct,decreasing = TRUE)
par(mar=c(11,4,4,4));barplot(sort(percAct,decreasing = TRUE),las=2)
```

How many other activies musicians are involved in & does this differ from other activities?
```{r}
m <- data.matrix(actData[actData$music_p12==1 & !is.na(actData$music_p12), paste0(activities,"_p12")])
otherActivities = rowSums(m,na.rm=TRUE) - 1
par(mar=c(11,4,4,4));hist(otherActivities, breaks=0:20, freq=FALSE)

means=list()
for (act in activities) {
  col = paste0(act,"_p12");
  m <- data.matrix(actData[actData[col]==1 & !is.na(actData[col]), paste0(activities,"_p12")])
  otherActivities = rowSums(m,na.rm=TRUE) - 1
  par(mar=c(11,4,4,4));hist(otherActivities, breaks=0:26, freq=FALSE,main=col)
  means[act] = mean(otherActivities, na.rm=TRUE)
}
means = as.numeric(means)
names(means)=activities

par(mar=c(11,4,4,4));barplot(sort(means,decreasing = TRUE),las=2)

```



Activity participation by site
```{r}
for (site in levels(actData$site_id_l)) {
  thisSite = actData$site_id_l==site
  percp12 <- 100*sapply(actData[thisSite, paste0(activities,"_p12")],sum,na.rm=TRUE) / sum(thisSite)
  par(mar=c(11,4,4,4));barplot(sort(percp12,decreasing = TRUE),las=2,main=site)
}
```

Plot correlation between activities, hierarchically clustered
# Plot activities correlation
```{r}
#analyze correlations across different activities
mat <- data.matrix(actData[paste0(activities,"_p12")])
mat <- data.matrix(actData[paste0(activities,"_any")])

mat[is.na(mat)] = 0
actCor <- cor(mat)
actCor <- actCor - diag(nrow(actCor))
df = 11000
t = actCor / sqrt( (1 - actCor^2) / (11000) )
p.value = 1 - pt(abs(t),df)
#emphasize colors (limits don't work properly on corrplot--cl.lim should give fullscale color, but instead only censors)
actCor <-  actCor * 2
lim <- 1
actCor[actCor > lim] = lim
actCor[actCor < -lim] = -lim


#corrplot(actCor, method="square",order="alphabet",diag=FALSE,cl.lim=c(-lim, lim))
corrplot(actCor, method="square",order="hclust",diag=FALSE,addrect=10,cl.lim=c(-lim, lim),p.mat = p.value, sig.level=0.05, insig="blank",tl.cex=.7)
```

Show similarity in a network form
https://psych-networks.com/r-tutorial-identify-communities-items-networks/
```{r}
library(qgraph)
cc=cor_auto(mat)
#corrplot(cc, method="square",order="hclust",diag=FALSE,addrect=6,p.mat = p.value, sig.level=0.05, insig="blank",tl.cex=.7)

graph<-qgraph(cc, graph="glasso", layout="spring", sampleSize = nrow(mat),
              vsize=7, cut=0, maximum=.45, border.width=1.5,threshold=TRUE)

library("igraph")
g = as.igraph(graph, attributes=TRUE)
sgc <- spinglass.community(g)
sgc$membership

#graph1<-qgraph(cc, graph="glasso", layout="spring", sampleSize = nrow(mat),
#              vsize=7, cut=0, maximum=.45, border.width=1.5,threshold=TRUE,
#              groups = sgc)


```

Do the correlation separately for each site
```{r}
for (site in levels(actData$site_id_l)) {
  mat <- data.matrix(actData[actData$site_id_l==site, paste0(activities,"_p12")])
  mat[is.na(mat)] = 0
  actCor <- cor(mat)
  #emphasize colors (limits don't work properly on corrplot--cl.lim should give fullscale color, but instead only censors)
  actCor <-  actCor * 2
  actCor[is.na(actCor)] = 0
  lim <- 1
  actCor[actCor > lim] = lim
  actCor[actCor < -lim] = -lim
  #corrplot(actCor, method="square",order="alphabet",diag=FALSE,cl.lim=c(-lim, lim))
  corrplot(actCor, method="square",order="hclust",diag=FALSE,addrect=10,cl.lim=c(-lim, lim),tl.cex=0.75)
}
```

analyze location of different activities
```{r}
#https://stackoverflow.com/questions/43624391/r-dplyr-methods-inside-own-function
summary.by <- function(data,var,group) {
  var <- enquo(var)
}

#% by region
regional <- actData %>% group_by(site_id_l) %>% mutate(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE))
print('=== participation by region ===')
print(arrange(aggregate(cbind(siteN,musicN, musicPerc) ~ site_id_l, data=regional, FUN=mean), desc(musicPerc)), row.names=FALSE)

#by gender
gender <- actData %>% group_by(sex) %>% mutate(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE)) #needed?
gender <- gender %>% group_by(sex) %>% summarise(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE))

#remove non-binary
#gender <- filter(gender, sex)
print('=== participation by gender ===')
print(arrange(aggregate(cbind(siteN,musicN, musicPerc) ~ sex, data=gender, FUN=mean), row.names=FALSE))

# participation.summary <- function(activity, data) {
#   fn <- paste0("sai_p_",activity,"_",c("school","outside","private","self"))
#
#   return(any(data[2,fn]))
# }
#
# unlist(lapply(activities, participation.summary(x,data)))

#setnames(mdata,old=c("school","outside"),new=c("Group_school","Group_outside"))
```

# TODO: total hours music, how well predicted by many other variables?

How much extra variance does intensity of different activities predict?

```{r echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide'}

source('/Users/jri/Documents/ Research/Projects/simphony/ABCD/Code/cmig_library/univariate_models.r')

phenoVars <- names(actData)[grepl('nihtbx',names(actData))]
neurocogVarPc <- names(actData)[grepl('neurocog',names(actData))]
rsVars <- names(actData)[grepl('rsfmri',names(actData))]
demoVars <- c("src_subject_id","eventname","abcd_site","rel_family_id","rel_group_id","age", "sex","household.income","high.educ","anthro_bmi_calc","race_ethnicity",'EUR', 'AFR', 'EAS', 'AMR',"accult_phenx_q1_p","accult_phenx_q1")

# TODO: family repeated structure
random_effects='~(1|abcd_site/rel_family_id)'

# Define Models
pc_cols = paste0('PC', seq(10))
covariates = c('age', 'sex',  pc_cols, 'household.income','high.educ')
random_effects='~(1|abcd_site/rel_family_id)'
IVs <- c("arts_hrlifetime","arts_hrperwk","sports_hrlifetime","sports_hrperwk","music_hrlifetime","music_hrperwk","soc_hrlifetime","soc_hrperwk","dance_hrlifetime","dance_hrperwk")
IVs <- c("arts_hrlifetime","sports_hrlifetime","music_hrlifetime","soc_hrlifetime")
DVs <- c('nihtbx_fluidcomp_uncorrected', 'nihtbx_cryst_uncorrected','EducAttain','IQ') 
#DVs <- names(actData)[grepl("nihtbx",names(actData))]
#DVs <- names(actData)[grepl("rsfmri",names(actData))]
#DVs <- names(actData)[grepl("neurocog",names(actData))]
#DVs <- c('EducAttain') #,'ASD','ADHD')

DVs <- c("arts_hrlifetime","sports_hrlifetime")
IVs <- c("nihtbx_cryst_uncorrected")

#swap DV/IV
#tmp = DVs
#DVs = IVs
#IVs = tmp

# TODO move this into run_model, for each IV
ad = actData[,c(IVs,DVs,covariates,"abcd_site","rel_family_id")]
# add more vars to gamma distribution list if plan to use as DV, or use quantnorm version as DV

model_table = create_model_table(DVs, IVs, covariates, random_effects, reduced_models=TRUE)

model_table_gamm = run_models(model_table, ad, model_type='GAMM4', cores=1)

#model_table_glm = run_models(model_table, ad, model_type='glm', cores=1)

# result = cbind(model_table_gamm$DV, model_table_gamm$DV, model_table_gamm$pval, model_table_gamm$r2_delta, model_table_gamm$r2_full)
# print(result)
```
 
# display fit results
```{r}
#data are in long form, indexed by IV and DV, use r2_delta as value

library(ggplot2)
library(patchwork)

plt1 <- ggplot(model_table_gamm, aes(DV, IV)) +
  geom_tile(aes(fill = r2_delta*100)) +
  geom_text(aes(label = round(r2_delta*100, 2))) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100*Delta~R^2)) 

plt2 <- ggplot(model_table_gamm, aes(DV, IV)) +
  geom_tile(aes(fill = 100*r2_full)) +
  geom_text(aes(label = round(100 * r2_full, 1))) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100 * R^2)) 

plt3 <- ggplot(model_table_gamm, aes(DV, IV)) +
  geom_tile(aes(fill = 100*r2_delta/r2_full)) +
  geom_text(aes(label = round(100*r2_delta/r2_full, 1))) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100*Delta~R^2 / R^2)) 

plt4 <- ggplot(model_table_gamm, aes(DV, IV)) +
  geom_tile(aes(fill = coeff)) +
  geom_text(aes(label = round(coeff, 3))) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_distiller(palette = "RdBu", name = expression(beta),limits = c(-max(abs(model_table_gamm$coeff)), max(abs(model_table_gamm$coeff))))

plt5 <- ggplot(model_table_gamm, aes(DV, IV)) +
  geom_tile(aes(fill = -log10(pval))) +
  geom_text(aes(label = round(-log10(pval), 1))) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(-log10(p)))  

pl <- align_patches(plt1, plt2, plt3, plt4, plt5)
pl[[1]]
pl[[2]]
pl[[3]] 
pl[[4]]
pl[[5]]



```


# The Following is ??
#save for aggregation-------------------------------------------------------
shortData <- mdata[,c("src_subject_id", "anyArt", "p12")]
setnames(shortData, old=c("anyArt", "p12"), new=c(theArt, paste0(theArt,"_p12")))
if (artIdx==1) {
  allData <- shortData
} else {
  allData = full_join(allData, shortData)
}

# Descriptive analysis-------------------------------------------------------

print(paste("======== ", theArt, " ========"))

#%music overall
mdata3 <- mdata %>% mutate(percArt = 100 * sum(anyArt,na.rm=TRUE)/n(), percArtP12= 100 * sum(p12,na.rm=TRUE)/n(), artN = sum(anyArt,na.rm=TRUE), siteN = n())
print('=== participation ===')
print(paste("N=",mdata3[1,"siteN"], ", artN=", mdata3[1,"artN"] , ", percent=", mdata3[1,"percArt"], ", active=", mdata3[1,"percArtP12"]))

# music in past 12 months (assumes music at some point in life--question only asked if they indiate some music)
table(mdata[,"p12"],exclude = NULL)
#   0    1 <NA>?
# 255 1656 2613
# implies 1656+255 = 1911 have some music ---42% of total 4524

# relationships between four types of music activity
table(mdata[,c("Group_school","Group_outside")])
table(mdata[,c("private","self")])

table(mdata[,c("Group_school","Group_outside","private","self")])

cor(mdata[,c("Group_school","Group_outside","private","self")], use = "complete.obs")

#mean practice amounts
onlymdata <- subset(mdata,anyArt==TRUE)
print('=== mean practice amounts ===')
print(summary(onlymdata[,c("nyr","nmonth","perwk","tspent")]))

#my.f = function(x) c(mean = mean(x,na.rm=TRUE), median = median(x,na.rm=TRUE))
#onlymdata[, sapply(.SD, my.f), .SDcols = c("nyr","nmonth","perwk","tspent"), by = site_name]


#music % by region
mdata2 <- mdata %>% group_by(site_id_l) %>% mutate(percArt = 100 * sum(anyArt,na.rm=TRUE)/n(), artN = sum(anyArt,na.rm=TRUE), siteN = n())
print('=== participation by region ===')
print(arrange(aggregate(cbind(siteN,artN, percArt) ~ site_id_l, data=mdata2, FUN=mean), desc(percArt)), row.names=FALSE)


# summary of demographics
mdata$sex = factor(mdata$sex)
print(summary(mdata[,c("ageYrs", "sex")]))

#annalyze aggregation
if (FALSE) {
  allData$anyArt = allData$music | allData$dance | allData$art | allData$drama | allData$crafts
  allData$anyArt_p12 = allData$music_p12 | allData$dance_p12 | allData$art_p12 | allData$drama_p12 | allData$crafts_p12
  allData$nArt = rowSums(allData[,arts], na.rm=TRUE)
  allData$nArt_p12 = rowSums(allData[,paste0(arts,"_p12")], na.rm=TRUE)s
  

  print(table(allData[,"nArt"],exclude = NULL))
  print(table(allData[,"nArt_p12"],exclude = NULL))

  print(paste("N art=",sum(allData$anyArt, na.rm=TRUE), ", N art p12=", sum(allData$anyArt_p12, na.rm=TRUE)))


}

