---
title: "ABCD 5.1 activities questionnaire"
author: "John R Iversen"
output: github_document
fig_width: 6 
fig_height: 4 
---

Date Prepared: `r Sys.time()`
Analyze ABCD Activities questionnaire from ABCD 5.1 (doi:10.15154/z563‑zd24). 

Run after ph_p_saiq_5.1.R

Updates: now has full DEAP demographics

NB: can extract all R code into a new file using
knitr:purl("/Users/jri/Documents/ Research/Projects/simphony/ABCD/Code/R/10_AnalyzeActivities_all_4.1.Rmd",documentation=2)
It can be easier to debug


# Preamble
```{r setup-chunk, include=FALSE}
knitr::opts_chunk$set(dev = "jpeg",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)
```



```{r load-libraries, results="hide"}
rm(list=ls())

doSaveFig = FALSE

#library(dtplyr)
library(dplyr)
library(tidyr)
library(data.table)
library(gsubfn)
library(rlang)
library(corrplot)
library(tibble)
library(ggplot2)
library(gamm4)
library(MuMIn)
```

# Load Data from ph_p_saiq_5.1.R, if needed
```{r load-data}

# have to define the timestamp of data we wish to use
ts <- "2023-12-02_16-07-06"

source("./setupPaths_5.1.R")

resultDir = file.path(resultDirRoot,ts)
figDir = file.path(resultDir,'FIGS')

if (!"actData" %in% ls()) {
  activitiesFile      <- file.path(resultDir,"abcd_5.1_activities.Rds")
  #activitiesFileLong  <- file.path(resultDir, "nda2.0.1_activities_long.Rds")
  #activitiesFileMeta  <- file.path(resultDir, "nda2.0.1_activities_meta.RData")
  activitiesFileMusic <- file.path(resultDir, "abcd_5.1_activities_music.csv")
  activitiesFileInst <- file.path(resultDir, "abcd_5.1_activities_inst.Rds")

  actData <- readRDS(activitiesFile)
  colnames(actData) <- gsub("sai_p_", "", colnames(actData)) #remove prefix from all var names
  actData0 = actData

  #actDataL <- readRDS(activitiesFileLong)
  #load(activitiesFileMeta)
  musicData <- fread(activitiesFileMusic)
  instData <- readRDS(activitiesFileInst)
  
  # strip prefix from all column names
  colnames(actData) <- gsub("sai_p_", "", colnames(actData))
}

```

In 5.x We have full activities data for baseline, year 1, year 2 and year 3; partial for year 4.

# Plot basic facts about each activity by timepoint

```{r activity-description-by-timepoint, fig.width=8, fig.height=6}

events = c('baseline_year_1_arm_1','1_year_follow_up_y_arm_1','2_year_follow_up_y_arm_1','3_year_follow_up_y_arm_1','4_year_follow_up_y_arm_1')
hrlim=5

activities_full <- unlist(strapplyc(names(actData),"(.*)_school$"))
measurements <- c("school",	"outside",	"private",	"self",	"nyr",	"nmonth",	"perwk",	"tspent",	"p12", "any", "ageStarted", "hrperwk","hrperyr", "hrlifetime","endorsed")

activities <- c('music')
for (act in activities) {
  
  actVars <- paste0(act,"_",measurements)
  
  for (event in events) {
    
    print(event)
    #mean age
    print(mean(actData[actData$eventname==event,'ageYrs']/12,na.rm=T))
    adata <- actData[actData$eventname==event, actVars]
    any <- adata[[paste0(act,'_endorsed')]] == TRUE
    adata <- adata[any,]
    n=sum(adata[[paste0(act,'_endorsed')]],na.rm=TRUE)
    print(n)
    
    if (doSaveFig) jpeg(file.path(figDir, paste0(act,'_participation@', event, ".jpeg")), units="in", width=8, height=6, res=300, quality=100)
  
    par(mfrow = c(2, 2))
    par(cex = 0.6)
    par(mar = c(3, 3, 1, 1), oma = c(1, 1, 1, 1))
    hist(adata[[paste0(act,'_ageStarted')]], main=paste0(act,' ageStarted'),xlim=c(0,15),freq=TRUE)
    hist(adata[[paste0(act,"_hrperwk")]], main=paste0(act,' Hrs/wk'),xlim=c(0, 8),freq=FALSE,breaks=seq(0,24*7,0.5))
    hist(adata[[paste0(act,"_hrperyr")]], main=paste0(act,' Hrs/yr'),freq=FALSE,breaks=seq(0,5000,50),xlim=c(0, 500))
    hist(adata[[paste0(act,'_hrlifetime')]], main=paste0(act,' hr lifetime'),breaks = 50,freq=FALSE, xlim=c(0, 1500))
    
    if (doSaveFig) dev.off()
    

    print(summary(adata[[paste0(act,"_hrperwk")]]))
    print(100*sum(adata[[paste0(act,"_hrperwk")]] > hrlim, na.rm = T)/n)
    
    print(summary(adata[[paste0(act,"_hrlifetime")]]))
    
    #get percentage participating in each context
    n=sum(adata[[paste0(act,'_any')]],na.rm=TRUE)
    percContext = apply(adata[,1:4],2,sum,na.rm=TRUE) / n
    barplot(percContext,names.arg=c("school","outside","private","self"),main="Contexts")
    
    #correlation of contexts
    mat = data.matrix(adata[,1:4])
    mat[is.na(mat)]=0
    cc = cor(mat)
    corrplot(cc,diag=FALSE)
  }
}
  
```

# Trends in music engagement (hrperwk) across time
```{r trends-across-timepoint, fig.width=10, fig.height=6}
#define activity and outcome measure
act <- 'music'
measure <- 'hrperwk'
measVar <- paste0(act,"_",measure)
actVars <- paste0(act,"_",measurements)

adata <- actData[, c('src_subject_id','eventname','ageYrs',actVars)]

#rename events
adata[adata$eventname=='baseline_year_1_arm_1','tpname'] = 'BL (10)'
adata[adata$eventname=='1_year_follow_up_y_arm_1','tpname'] = 'Y1 (11)'
adata[adata$eventname=='2_year_follow_up_y_arm_1','tpname'] = 'Y2 (12)'
adata[adata$eventname=='3_year_follow_up_y_arm_1','tpname'] = 'Y3 (13)'
adata[adata$eventname=='4_year_follow_up_y_arm_1','tpname'] = 'Y4 (14)'

adata[!is.na(adata[[measVar]]) & adata[[measVar]]==0, measVar]=NA #convert 0 hours per weeek to 0s

summary(adata)

#ggplot(adata,aes(x=tpname, y=music_hrperwk)) + geom_violin(draw_quantiles = c(0.5))+ylim(0,hrlim)

ggplot(adata,aes(x=.data$tpname, y=.data[[measVar]])) + 
  geom_violin(draw_quantiles = c(0.5))+ylim(0,hrlim) +
  theme(text = element_text(size = 16)) 
#scale_fill_viridis(discrete=TRUE) +
#scale_color_viridis(discrete=TRUE) 

if (doSaveFig) ggsave(paste0(measVar,"__tp.png"), device="png", path=figDir, dpi=300)

ggplot(adata,aes(x=tpname,y=music_hrperwk))+geom_boxplot() + ylim(0,hrlim)


ggplot(adata,aes(x=music_hrperwk, fill=tpname)) +
  geom_histogram(data = subset(adata, eventname == 'baseline_year_1_arm_1'), position='identity',alpha=0.3,bins=10, aes(y=(..count..)/sum(..count..))) +
  
  geom_histogram(data = subset(adata, eventname == '1_year_follow_up_y_arm_1'), position='identity',alpha=0.3,bins=10, aes(y=(..count..)/sum(..count..))) +
  
  geom_histogram(data = subset(adata, eventname == '2_year_follow_up_y_arm_1'), position='identity',alpha=0.3,bins=10, aes(y=(..count..)/sum(..count..))) + 
  
  geom_histogram(data = subset(adata, eventname == '3_year_follow_up_y_arm_1'), position='identity',alpha=0.3,bins=10, aes(y=(..count..)/sum(..count..))) +
  
  geom_histogram(data = subset(adata, eventname == '4_year_follow_up_y_arm_1'), position='identity',alpha=0.3,bins=10, aes(y=(..count..)/sum(..count..))) +
  
  xlim(0,hrlim)
  
```

# spaghetti plots

``` {r spaghetti-plots, fig.width=10, fig.height=8}

activities_full <- unlist(strapplyc(names(actData),"(.*)_school$"))
measurements <- c("school",	"outside",	"private",	"self",	"nyr",	"nmonth",	"perwk",	"tspent",	"p12", "any", "ageStarted", "hrperwk","hrperyr", "hrlifetime","endorsed")

#define activity and outcome measure
act <- 'music'
measure <- 'hrlifetime'
measVar <- paste0(act,"_",measure)
endorsedVar <- paste0(act,"_","endorsed")
actVars <- paste0(act,"_",measurements)

adata <- actData[, c('src_subject_id','eventname','ageYrs', actVars)]

# rename events
adata[adata$eventname=='baseline_year_1_arm_1','tpname'] = 'BL'
adata[adata$eventname=='1_year_follow_up_y_arm_1','tpname'] = 'Y1'
adata[adata$eventname=='2_year_follow_up_y_arm_1','tpname'] = 'Y2'
adata[adata$eventname=='3_year_follow_up_y_arm_1','tpname'] = 'Y3'
adata[adata$eventname=='4_year_follow_up_y_arm_1','tpname'] = 'Y4'

events = c('BL','Y1','Y2','Y3','Y4')

# widen the data for our measure to use in subject selection
adataw <- adata %>% select(-eventname) %>% pivot_wider(names_from = tpname, names_sep='.', values_from = all_of(c("ageYrs", actVars)))

nTPendorsed <- apply(adataw[,paste0(endorsedVar,'.',events)], 1, sum,na.rm=TRUE)
nTPendorsed[nTPendorsed==0] = NA
musicSubj = adataw[nTPendorsed>0,'src_subject_id'][[1]] #[1:10]

adata <- adata[adata$src_subject_id %in% musicSubj,]

#cbind(adataw[,paste0(endorsedVar,'.',events)], nTPendorsed)

hist(nTPendorsed, breaks = c(.5, 1.5, 2.5, 3.5, 4.5, 5.5), xlab="Number of Timepoints", ylab="Number of participants", main=paste0(act, " participation endorsements")) #bars centered on integers

#function to create a spaghetti plot of change in variable y across timepoints
tpPlot <- function(x,y,doLog, title) {
  p <- adata %>%
    group_by(src_subject_id) %>%
    ggplot(aes(x=get(x),y=get(y),group=src_subject_id)) +
    geom_line(alpha=.05, lwd=.5) + 
    geom_point(aes(x=get(x), y=get(y)), size=0.1, alpha=0.02) +
    #stat_smooth(aes(group=1),method="loess") + 
    labs(title=title, y=y, x=x) +
    theme(plot.title=element_text(face="bold", size=15))
  #geom_jitter(width=0.2)

  if (doLog) p <- p + scale_y_log10(labels = function(x) format(x, drop0trailing=T, scientific = FALSE))
  
  p
}

x="tpname"
y=measVar
title=paste0("Change in ", y, " across ", x)
doLog = FALSE
tpPlot(x,y,doLog, title)

doLog = TRUE
tpPlot(x,y,doLog, title)

x="tpname"
y=paste0(act,"_","hrperwk")
title=paste0(y, " at each ", x)
doLog = FALSE
tpPlot(x,y,doLog, title)


```

# starts and stops, napoleon style

```{r napoleon, fig.width=10, fig.height=8}

#starts and stops (stop means if hrlifetime has not increased from one timepoint to the next_
# TODO: extend to all timepoints

activities_full <- unlist(strapplyc(names(actData),"(.*)_school$"))
measurements <- c("school",	"outside",	"private",	"self",	"nyr",	"nmonth",	"perwk",	"tspent",	"p12", "any", "ageStarted", "hrperwk","hrperyr", "hrlifetime","endorsed")

#define activity and outcome measure
act <- 'music'
measure <- 'hrlifetime'
measVar <- paste0(act,"_",measure)
endorsedVar <- paste0(act,"_","endorsed")
actVars <- paste0(act,"_",measurements)

adata <- actData[, c('src_subject_id','eventname','ageYrs', endorsedVar)]

# rename events
events = c('baseline_year_1_arm_1','1_year_follow_up_y_arm_1','2_year_follow_up_y_arm_1','3_year_follow_up_y_arm_1','4_year_follow_up_y_arm_1')
tpnames = c('Y0','Y1','Y2','Y3','Y4')
index = 0:4
for (ii in 1:length(events)) {
  adata[adata$eventname==events[ii],'tpname'] = tpnames[ii] #crazy discovery that if will create list columns in the wide data below if all names_from are NA!!
  #adata[adata$eventname==events[ii],'tpidx'] = index[ii]
}

# widen the data for our measure to use in subject selection
adataw <- adata %>% select(-eventname) %>% pivot_wider(names_from = tpname, names_sep='.', values_from = all_of(c("ageYrs", endorsedVar)))

#sort endorsement columns
endCols <- paste0(endorsedVar, '.Y', 0:3)

adataw <- adataw[ do.call(order, -adataw[endCols]), ]

#deal with NAs; BL any NA -> FALSE; for other cols, replace NA with the and of the column to the left and right.
adataw <- adataw %>%
  mutate(music_endorsed.Y1 = ifelse(is.na(music_endorsed.Y1) & !is.na(music_endorsed.Y0 & music_endorsed.Y2), music_endorsed.Y0 & music_endorsed.Y2, music_endorsed.Y1)) %>%
  mutate(music_endorsed.Y2 = ifelse(is.na(music_endorsed.Y2) & !is.na(music_endorsed.Y1 & music_endorsed.Y3), music_endorsed.Y1 & music_endorsed.Y3, music_endorsed.Y2)) %>%
  mutate(music_endorsed.Y3 = ifelse(is.na(music_endorsed.Y3) & !is.na(music_endorsed.Y2 & music_endorsed.Y4), music_endorsed.Y2 & music_endorsed.Y4, music_endorsed.Y3))

#any remaining NA -> FALSE
adataw[endCols][is.na(adataw[endCols])] <- FALSE

#adataw <- nest(adataw, names=all_of(endCols))
adataw$pattern <- apply(adataw[endCols], 1, paste, collapse=",")
adataw$patternl <- apply(adataw['pattern'], 1, function(x) {as.list(as.logical(unlist(strsplit(x, ","))))} )
adataw$patternl <- apply(adataw['pattern'], 1, function(x) {print(as.list(as.logical(unlist(strsplit(x, ",")))))} )

cadataw$firstmusic <- apply(adataw$pattern, 1, function(x) which(as.logical(unlist(strsplit(names(x, ",")))))[1] ) #this monstrosity picks up the first value of pattern that is true

patterns <- table(adataw$pattern)

#does not work on standardized, clearly.
Npat = nrow(patterns)
Nsubj = sum(patterns)

colors = c('darkred','goldenrod2','dodgerblue3','chartreuse3')

y1=0
plot(c(0,4.2),c(0,Nsubj),type='n',main="music participation",ylab="participant",xlab="event")

for (iR in Npat:1) {
  pat = as.logical(unlist(strsplit(names(patterns[iR]), ",")))
  a <- which(pat)
  if (!length(a)) next
  b <- rep(NA,length(a))
  b[a] <- a
  x <- rbind(b-1, b)
  y2 <- y1+patterns[iR]
  p <- patterns[iR] / Nsubj
  # yy1 <- rep(NA,length(a))
  # yy2 <- rep(NA,length(a))
  # yy1[a] <- y1
  # yy2[a] <- y2
  
  c = c(rbind(b-1, b))
  
  for (iC in a) {
    x1 <- x[1,iC]
    x2 <- x[2,iC]
    thisColor = colors[a[1]]
    rect(x1,y1,x2,y2,col=thisColor,lwd=0)
    
    text(4.15,y1+(y2-y1)/2,sprintf("%.1f%%", p*100),cex=0.8)
  }
  y1=y2
  
}

# plot per-timepoint N above at y=8000
for (iT in 0:3) {
  col = adataw[[paste0(endorsedVar, '.Y', iT)]]
  Ntp = sum(col,na.rm=T)
  text(iT+0.5,8000,sprintf("N=%d (%.1f%%)", Ntp, Ntp/Nsubj*100))
}

# # music @ baseline N=4471 (4416 with new joining method)
# nT1 = sum(dat$mus_T1>0, na.rm=T)
# 
# # music both timepoints
# iT1T2 = dat$mus_T1>0 & dat$mus_T2>0
# nBoth = sum(iT1T2, na.rm=T)
# nBoth
# 
# # start @ T2  N=1316
# iT2 = dat$mus_T1==0 & dat$mus_T2>0
# nStartT2 = sum(iT2, na.rm=TRUE)
# nStartT2
# 
# # any music N=5787 (sum of previous 2 check)
# #sum(dat$mus_T1>0 | dat$mus_T2>0, na.rm=TRUE)
# 
# # stop after T1 N=563
# iT1 = dat$mus_T1>0 & (dat$mus_T2 == dat$mus_T1)
# nStopT2 = sum(iT1, na.rm=TRUE)
# nStopT2
# 
# pAll = (nT1-nStopT2) / N
# pT1Only = nStopT2 / N
# pT2Only = nStartT2 / N
# 
# plot(c(0,2.2),c(0,N),type='n',main="music participation",ylab="participant",xlab="event (BL, T2)")
# rect(0,0,2,nT1-nStopT2,col="darkred", border="transparent")
# rect(0,nT1-nStopT2,1,nT1,col="goldenrod2",border="transparent")
# rect(1,nT1,2,nStartT2+nT1,col="dodgerblue3", border="transparent")
# rect(0,0,2,N,col="transparent",border="black")
# 
# text(2.1,(nT1-nStopT2)/2,format(pAll*100,digits=3))
# text(2.1,(nT1-nStopT2/2),format(pT1Only*100,digits=3))
# text(2.1,(nT1+nStartT2/2),format(pT2Only*100,digits=3))
# 
# }


```
# now, colorize spaghetti plots according to year started

```{r spaghetti-tricolore, fig.width=10, fig.height=8 }



p <- adata %>%
    group_by(src_subject_id) %>%
    ggplot(aes(x=get(x),y=get(y),group=src_subject_id)) +
    geom_line(alpha=.05, lwd=.5) + 
    geom_point(aes(x=get(x), y=get(y)), size=0.1, alpha=0.02) +
    #stat_smooth(aes(group=1),method="loess") + 
    labs(title=title, y=y, x=x) +
    theme(plot.title=element_text(face="bold", size=15))
  #geom_jitter(width=0.2)

  if (doLog) p <- p + scale_y_log10(labels = function(x) format(x, drop0trailing=T, scientific = FALSE))

```

# Trends in engagement (hrperwk) by SES

```{r trends-by-SES}

measurements <- c("school",	"outside",	"private",	"self",	"nyr",	"nmonth",	"perwk",	"tspent",	"p12", "any", "ageStarted", "hrperwk","hrperyr", "hrlifetime","endorsed")
#define activity and outcome measure
act <- 'music'
measure <- 'hrperwk'
measVar <- paste0(act,"_",measure)
endorsedVar <- paste0(act,"_","endorsed")
actVars <- paste0(act,"_",measurements)

hrlim=3
adata <- actData[, c('src_subject_id','eventname','ageYrs','sex',paste0('PC',1:10),'household.income','high.educ','abcd_site','rel_family_id',actVars)]

# rename events
adata[adata$eventname=='baseline_year_1_arm_1','tpname'] = 'BL'
adata[adata$eventname=='1_year_follow_up_y_arm_1','tpname'] = 'Y1'
adata[adata$eventname=='2_year_follow_up_y_arm_1','tpname'] = 'Y2'
adata[adata$eventname=='3_year_follow_up_y_arm_1','tpname'] = 'Y3'
adata[adata$eventname=='4_year_follow_up_y_arm_1','tpname'] = 'Y4'

adata[!is.na(adata[[measVar]]) & adata[[measVar]]==0, measVar]=NA

ggplot(adata,aes(x=household.income, y=.data[[measVar]])) + geom_violin(draw_quantiles = c(0.5))+ ylim(0,hrlim)+
  theme(text = element_text(size = 20)) 
#scale_fill_viridis(discrete=TRUE) +
#scale_color_viridis(discrete=TRUE) 

ggplot(adata,aes(x=high.educ,y=.data[[measVar]])) + geom_violin(draw_quantiles = c(0.5))+ ylim(0,hrlim)+
  theme(text = element_text(size = 20)) 

#ggplot(adata,aes(x=tpname,y=.data[[measVar]]))+geom_boxplot() 


#ggplot(adata,aes(x=.data[[measVar]], fill=tpname)) + geom_histogram(position='identity',alpha=0.3,bins=50)

cov='ageYrs+sex+PC1+PC2+PC3+PC4+PC5+PC5+PC7+PC8+PC9+PC10'
#IV='household.income'
IV='high.educ'
DV=measVar
form = paste0(DV,' ~ ',IV,'+',cov)
form.r = paste0(DV,' ~ ',cov)
random='~(1|abcd_site/rel_family_id)'

minc<-gamm4(formula=formula(form),random=formula(random), data=adata)
minc.r<-gamm4(formula=formula(form.r),random=formula(random), data=adata)

summary(minc$gam)
summary(minc.r$gam)
anova(minc$gam,minc.r$gam)

r.squaredLR(minc$mer, minc.r$mer)

# high.educ explains 1.9% more variance (an increase of 50%)
```


# Plot participation in activities
simple distribution of participants endorsing any type of a given activity ever and in past 12 months
```{r activity-endorsement, warning=TRUE, fig.width=10, fig.height=6 }
activities <- activities_full[1:29]
percAct <- 100*sapply(actData[paste0(activities,"_any")],sum,na.rm=TRUE) / nrow(actData)
names(percAct) <- gsub("_any", "", names(percAct))

percp12 <- 100*sapply(actData[paste0(activities,"_p12")],sum,na.rm=TRUE) / nrow(actData)
names(percp12) <- gsub("_p12", "", names(percp12))


percSport <- 100*sum(actData$sports_any>0,na.rm=TRUE) / nrow(actData)
percSport_p12 <- 100*sum(actData$sports_p12>0,na.rm=TRUE) / nrow(actData)
percArts <- 100*sum(actData$arts_any,na.rm=TRUE) / nrow(actData)
percArts_p12 <- 100*sum(actData$arts_p12,na.rm=TRUE) / nrow(actData)
#percMusicNoSport <- 100*sum(actData$musicNoSport,na.rm=TRUE) / nrow(actData)
#percMusicNoSport_p12 <- 100*sum(actData$musicNoSport_p12,na.rm=TRUE) / nrow(actData)
aggPerc <- as.data.frame(cbind(percArts, percArts_p12, percSport, percSport_p12))

print("=== Activity endorsement: p12")
sort(percp12,decreasing = TRUE)
par(mar=c(11,4,4,4));barplot(sort(percp12,decreasing = TRUE), las=2, ylim=range(pretty(c(0, percp12))),ylab='Participation (%)',main='Participation by Activity')

print("=== Activity endorsement: Lifetime")
sort(percAct,decreasing = TRUE)
par(mar=c(11,4,4,4));barplot(sort(percAct,decreasing = TRUE), las=2, ylim=range(pretty(c(0, percAct))),ylab='Participation (%)',main='Participation by Activity')
```

##How many other activies musicians are involved in & does this differ from other activities?
```{r other-activities-by-activity}
m <- data.matrix(actData[actData$music_p12==1 & !is.na(actData$music_p12), paste0(activities,"_p12")])
otherActivities = rowSums(m,na.rm=TRUE) - 1
par(mar=c(11,4,4,4));hist(otherActivities, breaks=0:20, freq=FALSE)

means=list()
for (act in activities) {
  col = paste0(act,"_p12");
  m <- data.matrix(actData[actData[col]==1 & !is.na(actData[col]), paste0(activities,"_p12")])
  otherActivities = rowSums(m,na.rm=TRUE) - 1
  par(mar=c(11,4,4,4));hist(otherActivities, breaks=0:26, freq=FALSE,main=col)
  means[act] = mean(otherActivities, na.rm=TRUE)
}
means = as.numeric(means)
names(means)=activities

par(mar=c(11,4,4,4));barplot(sort(means,decreasing = TRUE), las=2, ylim=range(pretty(c(0, means))))

```



##Activity participation by site
```{r activity-by-site, eval=TRUE, echo=TRUE}

source("./ABCD_translation_tables_5.1.R")

activities <- activities_full[1:29]
for (siteid in levels(as.factor(actData$abcd_site))) {
  thisSite = actData$abcd_site==siteid
  percp12 <- 100*sapply(actData[thisSite, paste0(activities,"_p12")],sum,na.rm=TRUE) / sum(thisSite)
  par(mar=c(11,4,4,4));barplot(sort(percp12,decreasing = TRUE),las=2,main=siteTable[siteid])
}
```

Plot correlation between activities, hierarchically clustered
# Plot activities correlation
```{r activities-correlation, fig.width=10, fig.height=10}
#analyze correlations across different activities
mat <- data.matrix(actData[paste0(activities_full[1:29],"_p12")])
#mat <- data.matrix(actData[paste0(activities_full[1:29],"_any")])

# clean up labels
colnames(mat) <- gsub("_any", "", colnames(mat))
colnames(mat) <- gsub("_p12", "", colnames(mat))
colnames(mat) <- gsub("wpolo", "swim", colnames(mat))
colnames(mat) <- gsub("polo", "horse", colnames(mat))



mat[is.na(mat)] = 0
actCor <- cor(mat)
actCor <- actCor - diag(nrow(actCor))
df = 11000
t = actCor / sqrt( (1 - actCor^2) / (11000) )
p.value = 1 - pt(abs(t),df)
#emphasize colors (limits don't work properly on corrplot--cl.lim should give fullscale color, but instead only censors)
actCor <-  actCor * 3
lim <- 1
actCor[actCor > lim] = lim
actCor[actCor < -lim] = -lim

#corrplot(actCor, method="square",order="alphabet",diag=FALSE,cl.lim=c(-lim, lim))
corrplot(actCor, method="square",order="hclust",diag=FALSE,addrect=11,cl.lim=c(-lim, lim),p.mat = p.value, sig.level=0.05, insig="blank",tl.cex=1.5,col=colorRampPalette(c("blue","white","red"))(200))
```

Show similarity in a network form
https://psych-networks.com/r-tutorial-identify-communities-items-networks/
```{r activities-network, fig.width=10, fig.height=10}
library(qgraph)
cc=cor_auto(mat)
#corrplot(cc, method="square",order="hclust",diag=FALSE,addrect=6,p.mat = p.value, sig.level=0.05, insig="blank",tl.cex=.7)

graph<-qgraph(cc, graph="glasso", layout="spring", sampleSize = nrow(mat),
              vsize=7, cut=0, maximum=.45, border.width=1.5,threshold=TRUE)

library("igraph")
g = as.igraph(graph, attributes=TRUE)
sgc <- spinglass.community(g)
sgc$membership

#graph1<-qgraph(cc, graph="glasso", layout="spring", sampleSize = nrow(mat),
#              vsize=7, cut=0, maximum=.45, border.width=1.5,threshold=TRUE,
#              groups = sgc)


```

Do the correlation separately for each site
```{r activities-correlation-by-site, eval=FALSE, echo=FALSE}
for (site in levels(actData$abcd_site)) {
  mat <- data.matrix(actData[actData$abcd_site==site, paste0(activities,"_p12")])
  mat[is.na(mat)] = 0
  actCor <- cor(mat)
  #emphasize colors (limits don't work properly on corrplot--cl.lim should give fullscale color, but instead only censors)
  actCor <-  actCor * 2
  actCor[is.na(actCor)] = 0
  lim <- 1
  actCor[actCor > lim] = lim
  actCor[actCor < -lim] = -lim
  #corrplot(actCor, method="square",order="alphabet",diag=FALSE,cl.lim=c(-lim, lim))
  corrplot(actCor, method="square",order="hclust",diag=FALSE,addrect=10,cl.lim=c(-lim, lim),tl.cex=0.75)
}
```

analyze location of different activities
```{r activity-by-site-and-gender, eval=FALSE}
#https://stackoverflow.com/questions/43624391/r-dplyr-methods-inside-own-function
summary.by <- function(data,var,group) {
  var <- enquo(var)
}

#% by region
regional <- actData %>% group_by(abcd_site) %>% dplyr::mutate(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE))
print('=== participation by region ===')
print(arrange(aggregate(cbind(siteN,musicN, musicPerc) ~ abcd_site, data=regional, FUN=mean), desc(musicPerc)), row.names=FALSE)

#5.x range from 52% to 19% by site


#by gender
gender <- actData %>% group_by(sex) %>% dplyr::mutate(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE)) #needed?
gender_summary <- gender %>% group_by(sex) %>% dplyr::summarise(siteN = n(), musicPerc = 100 * sum(music_p12,na.rm=TRUE)/n(), musicN = sum(music_p12,na.rm=TRUE))

#remove non-binary
#gender <- filter(gender, sex)
print('=== participation by gender ===')
#print(arrange(aggregate(cbind(siteN, musicN, musicPerc) ~ sex, data=gender, FUN=mean), row.names=FALSE)) #another way to aggregate the data
print(gender_summary)

# participation.summary <- function(activity, data) {
#   fn <- paste0("sai_p_",activity,"_",c("school","outside","private","self"))
#
#   return(any(data[2,fn]))
# }
#
# unlist(lapply(activities, participation.summary(x,data)))

#setnames(mdata,old=c("school","outside"),new=c("Group_school","Group_outside"))
```

# total hours music, how well predicted by many other variables?

How much extra variance does intensity of different activities predict?

```{r modeling, eval=FALSE, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide'}

source('./nc_y_nihtb_5.1.R')
#source("./abcd_y_lt_5.1.R")

keyCols = c('src_subject_id','eventname')

actData = actData0
#actData <- left_join(actData, abcd_y_lt[,c(keyCols, 'rel_family_id')], by=keyCols, relationship="one-to-one")
actData <- left_join(actData, nc_y_nihtb[,c(keyCols, 'nihtbx_fluidcomp_uncorrected', 'nihtbx_cryst_uncorrected')], by=keyCols, relationship="one-to-one")


source('/Users/jri/Documents/matlab/matlab/projects/simphony/ABCD/cmig-research-group/cmig_library/univariate_models.r')

#phenoVars <- names(actData)[grepl('nihtbx',names(actData))]
#neurocogVarPc <- names(actData)[grepl('neurocog',names(actData))]
#rsVars <- names(actData)[grepl('rsfmri',names(actData))]
#demoVars <- c("src_subject_id","eventname","abcd_site","rel_family_id","rel_group_id","age", "sex","household.income","high.educ","anthro_bmi_calc","race_ethnicity",'EUR', 'AFR', 'EAS', 'AMR',"accult_phenx_q1_p","accult_phenx_q1")

# Define Models
pc_cols = paste0('PC', seq(10))
covariates = c('ageYrs', 'sex',  pc_cols, 'household.income','high.educ')
random_effects='~(1|abcd_site/rel_family_id)'
IVs <- c("arts_hrlifetime","arts_hrperwk","sports_hrlifetime","sports_hrperwk","music_hrlifetime","music_hrperwk","soc_hrlifetime","soc_hrperwk","dance_hrlifetime","dance_hrperwk")
IVs <- c("arts_hrlifetime","sports_hrlifetime","music_hrlifetime","soc_hrlifetime")
DVs <- c('nihtbx_fluidcomp_uncorrected', 'nihtbx_cryst_uncorrected') #,'prs_EducAttain','prs_IQ') 
#DVs <- names(actData)[grepl("nihtbx",names(actData))]
#DVs <- names(actData)[grepl("rsfmri",names(actData))]
#DVs <- names(actData)[grepl("neurocog",names(actData))]
#DVs <- c('prs_EducAttain') #,'prs_ASD','prs_ADHD')

#DVs <- c("arts_hrlifetime","sports_hrlifetime")
#IVs <- c("nihtbx_cryst_uncorrected")

#swap DV/IV
if (FALSE) {
  print('* Swapping DV & IV')
  tmp = DVs
  DVs = IVs
  IVs = tmp
}

# TODO move this into run_model, for each IV

eventnames <- c('baseline_year_1_arm_1', '1_year_follow_up_y_arm_1', '2_year_follow_up_y_arm_1', '3_year_follow_up_y_arm_1')

# single timepoint
if (FALSE) {
  eventnames <- c('2_year_follow_up_y_arm_1')
  eventnames <- c('baseline_year_1_arm_1')
  random_effects='~(1|abcd_site)' #no subject-level random effect because cross-sectional
}

ad = actData[actData$eventname %in% eventnames, c(IVs,DVs,covariates,"abcd_site","rel_family_id")]

subj = actData[actData$eventname %in% eventnames, "src_subject_id"]
nSubj = length(unique(subj[[1]]))
print(sprintf('== N subj = %d ==',nSubj))

# add more vars to gamma distribution list if plan to use as DV, or use quantnorm version as DV

model_table = create_model_table(DVs, IVs, covariates, random_effects, reduced_models=TRUE)

model_table_gamm = run_models(model_table, ad, model_type='GAMM4', cores=1,QN_IVs=TRUE,QN_gauss_DVs=TRUE)

#model_table_glm = run_models(model_table, ad, model_type='glm', cores=1)

# result = cbind(model_table_gamm$DV, model_table_gamm$DV, model_table_gamm$pval, model_table_gamm$r2_delta, model_table_gamm$r2_full)
# print(result)
```
 
# display fit results
```{r model-results, eval=FALSE, echo=FALSE, , fig.width=10, fig.height=5}
#data are in long form, indexed by IV and DV, use r2_delta as value

library(ggplot2)
library(patchwork)

tsize = 8
nudge = .005

plt1 <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = r2_delta*100)) +
  geom_text(aes(label = round(r2_delta*100, 2)), size=tsize, color="black") +
  geom_text(aes(label = round(r2_delta*100, 2)), size=tsize, color="white", nudge_x=-nudge, nudge_y=nudge) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100*Delta~R^2)) 

plt2 <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = 100*r2_full)) +
  geom_text(aes(label = round(100 * r2_full, 1)), size=tsize) +
  geom_text(aes(label = round(100 * r2_full, 1)), size=tsize, color="white", nudge_x=-nudge, nudge_y=nudge) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100 * R^2)) 

plt3 <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = 100*r2_delta/r2_full)) +
  geom_text(aes(label = round(100*r2_delta/r2_full, 1)), size=tsize) +
  geom_text(aes(label = round(100*r2_delta/r2_full, 1)), size=tsize, color="white", nudge_x=-nudge, nudge_y=nudge) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(100*Delta~R^2 / R^2)) 

plt4 <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = coeff)) +
  geom_text(aes(label = round(coeff, 3)), size=tsize) +
  geom_text(aes(label = round(coeff, 3)), size=tsize, color="white", nudge_x=-nudge, nudge_y=nudge) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_distiller(palette = "RdBu", name = expression(beta),limits = c(-max(abs(model_table_gamm$coeff)), max(abs(model_table_gamm$coeff))))

plt4b <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = coeff)) +
  geom_text(aes(label = round(coeff, 3)), size=tsize) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_distiller(palette = "RdBu", name = expression(beta),limits = c(-max(abs(model_table_gamm$coeff)), max(abs(model_table_gamm$coeff))))

plt5 <- ggplot(model_table_gamm, aes(IV, DV)) +
  geom_tile(aes(fill = -log10(pval))) +
  geom_text(aes(label = round(-log10(pval), 1)), size=tsize) +
  geom_text(aes(label = round(-log10(pval), 1)), size=tsize, color="white", nudge_x=-nudge, nudge_y=nudge) +
  theme(axis.text.x = element_text(angle = -45, hjust=0, vjust=1)) +
  scale_fill_continuous(name = expression(-log10(p)))  

pl <- align_patches(plt1, plt2, plt3, plt4, plt4b, plt5)
pl[[1]]
pl[[2]]
pl[[3]] 
pl[[4]]
pl[[5]]
pl[[6]]




```

``` {asis, echo=FALSE}
# The Following is ??
#save for aggregation-------------------------------------------------------
shortData <- mdata[,c("src_subject_id", "anyArt", "p12")]
setnames(shortData, old=c("anyArt", "p12"), new=c(theArt, paste0(theArt,"_p12")))
if (artIdx==1) {
  allData <- shortData
} else {
  allData = full_join(allData, shortData)
}

# Descriptive analysis-------------------------------------------------------

print(paste("======== ", theArt, " ========"))

#%music overall
mdata3 <- mdata %>% mutate(percArt = 100 * sum(anyArt,na.rm=TRUE)/n(), percArtP12= 100 * sum(p12,na.rm=TRUE)/n(), artN = sum(anyArt,na.rm=TRUE), siteN = n())
print('=== participation ===')
print(paste("N=",mdata3[1,"siteN"], ", artN=", mdata3[1,"artN"] , ", percent=", mdata3[1,"percArt"], ", active=", mdata3[1,"percArtP12"]))

# music in past 12 months (assumes music at some point in life--question only asked if they indiate some music)
table(mdata[,"p12"],exclude = NULL)
#   0    1 <NA>?
# 255 1656 2613
# implies 1656+255 = 1911 have some music ---42% of total 4524

# relationships between four types of music activity
table(mdata[,c("Group_school","Group_outside")])
table(mdata[,c("private","self")])

table(mdata[,c("Group_school","Group_outside","private","self")])

cor(mdata[,c("Group_school","Group_outside","private","self")], use = "complete.obs")

#mean practice amounts
onlymdata <- subset(mdata,anyArt==TRUE)
print('=== mean practice amounts ===')
print(summary(onlymdata[,c("nyr","nmonth","perwk","tspent")]))

#my.f = function(x) c(mean = mean(x,na.rm=TRUE), median = median(x,na.rm=TRUE))
#onlymdata[, sapply(.SD, my.f), .SDcols = c("nyr","nmonth","perwk","tspent"), by = site_name]


#music % by region
mdata2 <- mdata %>% group_by(abcd_site) %>% mutate(percArt = 100 * sum(anyArt,na.rm=TRUE)/n(), artN = sum(anyArt,na.rm=TRUE), siteN = n())
print('=== participation by region ===')
print(arrange(aggregate(cbind(siteN,artN, percArt) ~ abcd_site, data=mdata2, FUN=mean), desc(percArt)), row.names=FALSE)


# summary of demographics
mdata$sex = factor(mdata$sex)
print(summary(mdata[,c("ageYrs", "sex")]))

#annalyze aggregation
if (FALSE) {
  allData$anyArt = allData$music | allData$dance | allData$art | allData$drama | allData$crafts
  allData$anyArt_p12 = allData$music_p12 | allData$dance_p12 | allData$art_p12 | allData$drama_p12 | allData$crafts_p12
  allData$nArt = rowSums(allData[,arts], na.rm=TRUE)
  allData$nArt_p12 = rowSums(allData[,paste0(arts,"_p12")], na.rm=TRUE)s
  

  print(table(allData[,"nArt"],exclude = NULL))
  print(table(allData[,"nArt_p12"],exclude = NULL))

  print(paste("N art=",sum(allData$anyArt, na.rm=TRUE), ", N art p12=", sum(allData$anyArt_p12, na.rm=TRUE)))


}
```
