---
title: "Bivariate latent change score: music education and crystallized intelligence"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```


Define variables
```{r}
rm(list=ls())

#define activity variable of interest
act = 'music'
#act = 'sports'
actVar = paste0(act, '_hrlifetime')

#define cognitive variable of interest
cogs = c("picvocab", "flanker","list","cardsort","pattern","picture","reading","fluidcomp","cryst","totalcomp")
idx = 8 #fluid 9 #cryst
cog = cogs[idx]
cogVar = paste0("nihtbx_", cog,"_uncorrected")

#cog = "smri_area_cort.fuzzy12_central.lh"
#cog = "smri_area_cort.fuzzy12_occipital.lh"

#cog = "dmri_dti.vol_fiber.at_tslf.rh"
```

Load
```{r}
library(lavaan)
library(ggplot2)
library(reshape2)

#=================
#==  load data  ==
#=================

ts <- "2023-12-02_12-21-34"

source("./setupPaths_5.1.R")
# I/O

resultDir = file.path(resultDirRoot,ts)

activitiesFile <- file.path(resultDir,"abcd_5.1_activities.Rds")

actData <- readRDS(activitiesFile)

source('./nc_y_nihtb_5.1.R')

keyCols = c('src_subject_id','eventname')

actData <- left_join(actData, nc_y_nihtb[,c(keyCols, 'nihtbx_fluidcomp_uncorrected', 'nihtbx_cryst_uncorrected')], by=keyCols, relationship="one-to-one")

dat0 = actData


```

```{r}

# include only baseline and year 2 (TODO: There is activity data at year 1 and 3 as well)

saiqEventnames <- c("baseline_year_1_arm_1","2_year_follow_up_y_arm_1")

dat = dat0[,c("src_subject_id","eventname",actVar,cogVar)]
dat = dat[dat$eventname %in% saiqEventnames,]
#save copy into "dat_main"
dat_main = dat

#create selection vectors for each timepoint, use to look up 'previous' timepoint's value for nyr
tpidx <- list()
en <- dat[['eventname']]
for (i in 1:length(saiqEventnames)) {
  tpidx[i] <- list(grepl(saiqEventnames[i], en))
}
names(tpidx) <- saiqEventnames

#Split into columns by timepoint
dat_all = reshape()

#=================
#==  analysis  ==
#=================


#========================================================
# ---- use standardized measures from here out ----
#========================================================
# datfu2 = datfu
# names(datfu2) = gsub("_l", "", names(datfu2))
# dat2 = rbind(datbl,datfu2)
# dat2 = process_dat(dat2, contains_l_vars = F)
# 
# dat = dat2
# dat$nihtbx_cryst_uncorrected = scale(dat$nihtbx_cryst_uncorrected)
# dat$sports_activity_ss_music_nmonth_p = scale(dat$sports_activity_ss_music_nmonth_p)
# 
# datbl = dat[dat$eventname == "baseline_year_1_arm_1",]
# datfu = dat[dat$eventname == "2_year_follow_up_y_arm_1",]
# names(datfu)[3:6] = paste0(names(datfu)[3:6], "_l")
# datbl$eventname = NULL
# datfu$eventname = NULL
# 
# dat = merge(datbl,datfu, by = "src_subject_id", all = T)


dat$mus_T1 = dat[[actVar]]
dat$mus_T2 = dat[[paste0(actVar,"_l")]]

dat_all = dat #store with all participants

```
Look at patterns of music engagement:
```{r}
dat = dat_all
N = nrow(dat)

# music @ baseline N=4471
nT1 = sum(dat$mus_T1>0)

# music both timepoints
iT1T2 = dat$mus_T1>0 & dat$mus_T2>0
nBoth = sum(iT1T2, na.rm=T)
nBoth

# start @ T2  N=1316
iT2 = dat$mus_T1==0 & dat$mus_T2>0
nStartT2 = sum(iT2, na.rm=TRUE)
nStartT2

# any music N=5787 (sum of previous 2 check)
#sum(dat$mus_T1>0 | dat$mus_T2>0, na.rm=TRUE)

# stop after T1 N=563
iT1 = dat$mus_T1>0 & (dat$mus_T2 == dat$mus_T1)
nStopT2 = sum(iT1, na.rm=TRUE)
nStopT2

pAll = (nT1-nStopT2) / N
pT1Only = nStopT2 / N
pT2Only = nStartT2 / N

plot(c(0,2.2),c(0,N),type='n',main="music participation",ylab="participant",xlab="event (BL, T2)")
rect(0,0,2,nT1-nStopT2,col="darkred", border="transparent")
rect(0,nT1-nStopT2,1,nT1,col="goldenrod2",border="transparent")
rect(1,nT1,2,nStartT2+nT1,col="dodgerblue3", border="transparent")
rect(0,0,2,N,col="transparent",border="black")

text(2.1,(nT1-nStopT2)/2,format(pAll*100,digits=3))
text(2.1,(nT1-nStopT2/2),format(pT1Only*100,digits=3))
text(2.1,(nT1+nStartT2/2),format(pT2Only*100,digits=3))


```

# Various restrictions to subsets (pick one)
## include everyone with any music

```{r}
#include everyone with any music [N=5787]
dat=dat_all
dmus = dat$mus_T2-dat$mus_T1
summus = dat$mus_T2+dat$mus_T1
toInclude = (!is.na(dmus) & summus>0)
toInclude = dat$mus_T1>0 | dat$mus_T2>0
toInclude[is.na(toInclude)] = FALSE
N = sum(toInclude,na.rm=T)
dat = dat[toInclude,]

iT1T2 = iT1T2[toInclude]
iT1 = iT1[toInclude]
iT2 = iT2[toInclude]
```

## include everyone with music at both timepoints and is increasing
```{r}
#include everyone with music at both timepoints and is increasing [N=2066]
dat=dat_all

iInc = dat$mus_T1>0 & (dat$mus_T2 > dat$mus_T1)
toInclude = iInc
toInclude[is.na(toInclude)] = FALSE
N = sum(toInclude,na.rm=T)
dat = dat[toInclude,]

iT1T2 = iT1T2[toInclude] #verified all false
iT1 = iT1[toInclude] #verified all false
iT2 = iT2[toInclude] #verified all true
```


## include everyone with music at both timepoints and increasing or constant
```{r}
#include everyone with music at both timepoints and is increasing or constant [N=2629]
dat=dat_all

iInc = dat$mus_T1>0 & (dat$mus_T2 >= dat$mus_T1)
toInclude = iInc
toInclude[is.na(toInclude)] = FALSE
N = sum(toInclude,na.rm=T)
dat = dat[toInclude,]

iT1T2 = iT1T2[toInclude] #verified all false
iT1 = iT1[toInclude] #verified all false
iT2 = iT2[toInclude] #verified all true
```


## include only those starting at T2 (Fails)
```{r}
#include only those starting T2 [N=1316] *FAILS--no variance at T1, as all are 0
dat=dat_all

iInc = dat$mus_T1==0 & (dat$mus_T2 > 0)
toInclude = iInc
toInclude[is.na(toInclude)] = FALSE
N = sum(toInclude,na.rm=T)
dat = dat[toInclude,]

iT1T2 = iT1T2[toInclude]
iT1 = iT1[toInclude] #verified all false
iT2 = iT2[toInclude] #verified all false
```


## include only musicians with change in hours
```{r}
#include only musicians with change in hours [is there an age confound?] [N=3382]
dat = dat_all
dmus = dat$mus_T2-dat$mus_T1
toInclude = (dmus>0 & !is.na(dmus))
toInclude[is.na(toInclude)] = FALSE
dat = dat[toInclude==T,]

```

```{r}
#test those with no change (i.e. stopped music)--model doesn't work XX
dmus = dat$mus_T2-dat$mus_T1
dat = dat[dmus==0 & !is.na(dmus),]

```


```{r}
# log transform? 
logStr = "";
LOG = T

if (LOG){
  
  # as a convenience map 0 hours to 0 after log (makes plotting nicer, but need to undo when converting from log)
  dat$mus_T1[dat$mus_T1 == 0] = 1 #was "NA" but that loses anyone who started at tp2
  dat$mus_T2[dat$mus_T2 == 0] = 1
  
  dat$mus_T1 = log(dat$mus_T1)
  dat$mus_T2 = log(dat$mus_T2)
  
  logStr = "Log "

}

dat$cog_T1 = dat[[cogVar]]
dat$cog_T2 = dat[[paste0(cogVar,"_l")]]

```


Run Model
```{r}

#we use mus and cog as generics, but represent actual variable names in actVar and cogVar

BLCS<-'

mus_T2 ~ 1*mus_T1     # This parameter regresses MUS_T2 perfectly on MUS_T1
dmus1 =~ 1*mus_T2     # This defines the latent change score factor as measured perfectly by scores on MUS_T2
dmus1 ~ 1             # This estimates the intercept of the change score 
mus_T1 ~  1           # This estimates the intercept of MUS_T1 
mus_T2 ~ 0*1          # This constrains the intercept of MUS_T2 to 0

cog_T2 ~ 1*cog_T1     # This parameter regresses COG_T2 perfectly on COG_T1
dcog1 =~ 1*cog_T2     # This defines the latent change score factor as measured perfectly by scores on COG_T2
cog_T2 ~ 0*1          # This line constrains the intercept of COG_T2 to 0
cog_T2 ~~ 0*cog_T2    # This fixes the variance of the COG_T1 to 0  

dmus1 ~~  dmus1       # This estimates the variance of the change scores
mus_T1 ~~   mus_T1    # This estimates the variance of the MUS_T1 
mus_T2 ~~ 0*mus_T2    # This fixes the variance of the MUS_T2 to 0  

dcog1 ~ 1             # This estimates the intercept of the change score 
cog_T1 ~ 1            # This estimates the intercept of COG_T1 
dcog1 ~~ dcog1        # This estimates the variance of the change scores 
cog_T1 ~~ cog_T1      # This estimates the variance of COG_T1 

dcog1~mus_T1+cog_T1   # This estimates the MUS to COG coupling parameter and the COG to COG self-feedback
dmus1~cog_T1+mus_T1    # This estimates the COG to MUS coupling parameter and the NEU to NEU self-feedback

mus_T1 ~~  cog_T1     # This estimates the MUS_T1 COG_T1 covariance
dmus1~~dcog1          # This estimates the dMUS and dCOG covariance
'

cat(paste0("\n\n==== ", actVar ," vs. ", cogVar, " ====\n\n"))
fitBLCS <- lavaan(BLCS, data=dat, estimator='mlr',fixed.x=FALSE,missing='fiml',em.h1.iter.max=1000)
#robust ml, full information max likelihood (case-by-case with all available data)
summary(fitBLCS, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
varTable(fitBLCS)
```


Plotting

```{r}
library(lavaanPlot)

labels=list(mus_T1 = paste0(act,"_T1"), mus_T2 = paste0(act,"_T2"), dmus1 = paste0("d",act),cog_T1 = paste0(cog,"_T1"), cog_T2 = paste0(cog,"_T2"), dcog1 = paste0("d",cog))

lavaanPlot(name=cog,fitBLCS,coefs=T,covs=F,edge_options=list(color="grey"),labels=labels)
lavaanPlot(name=cog,fitBLCS,coefs=T,covs=F,edge_options=list(color="grey"),stand=T,labels=labels)


lavaanPlot(name=cog,fitBLCS,coefs=T,covs=T,stars="covs",edge_options=list(color="grey"),stand=F,labels=labels)

lavaanPlot(name=cog,fitBLCS,coefs=T,covs=T,stars="covs",edge_options=list(color="grey"),stand=T,labels=labels)

#stars: latent, regress, covs

labels=list(mus_T1 = paste0(act,"_T1"), mus_T2 = paste0(act,"_T2"), dmus1 = paste0("d",act),cog_T1 = paste0(cog,"_T1"), cog_T2 = paste0(cog,"_T2"), dcog1 = paste0("d",cog))
plot = lavaanPlot(name=cog,fitBLCS,coefs=T,covs=F,stars="covs",edge_options=list(color="grey"),labels=labels)
svg = DiagrammeRsvg::export_svg(plot)
cat(file=paste0(act,"_",cog,".svg"),svg)

```

```{r}

# summary(fitBLCS)
# cat("\n\n_____\n\n")
# summary(fitBLCS2)


# theme_set(theme_grey(base_size = 18)) #increase text size
samplesize = nrow(dat)
id=factor(1:samplesize)

musGroup = rep('',times=samplesize)
musGroup[iT1T2]='Both'
musGroup[iT1]='T1'
musGroup[iT2]='T2'

musGroupColor = rep('black',times=samplesize);
musGroupColor[iT1T2]='darkred'
musGroupColor[iT1]='goldenrod2'
musGroupColor[iT2]='dodgerblue3'

plotdattemp=data.frame(c(dat$mus_T1,dat$cog_T1),c(dat$mus_T2,dat$cog_T2),as.factor(c(id,id)),
                       c(rep('mus',times=samplesize),rep('cog',times=samplesize)),c(musGroup,rep('',times=samplesize)), c(musGroupColor,rep('',times=samplesize)))
colnames(plotdattemp)<-c('T1', 'T2','id','Domain','MusGroup','MusGroupColor')
plotdat<-melt(plotdattemp,by='id')

#ggplot(plotdat,aes(variable,value,group=id,col=Domain))+geom_point(size=1,alpha=.01)+geom_line(alpha=.01)+
#  ggtitle('Bivariate Latent Change Score model')+ylab('Scores')+xlab('Time points')+facet_grid(~Domain)

a=0.05

ggplot(plotdat[plotdat$Domain=="mus",],aes(variable,value,group=id))+geom_point(size=1,alpha=a)+geom_line(alpha=a)+
  ggtitle(paste0("Timepont changes in ",logStr,actVar))+ylab('Scores')+xlab('Time points')

p <- ggplot(plotdat[plotdat$Domain=="cog",],aes(variable,value,group=id))+geom_point(size=1,alpha=a)+geom_line(alpha=a)+
  ggtitle(paste0("Timepont changes in  ", cog))+ylab('Scores')+xlab('Time points')
ggExtra::ggMarginal(p,type="density",margins="y",size=4)

p <- ggplot(plotdat[plotdat$Domain=="mus",],aes(variable,value,group=id))+geom_point(size=1,alpha=a)+geom_line(alpha=a)+
  ggtitle(paste0("Timepont changes in ",logStr,actVar))+ylab('Scores')+xlab('Time points')
ggExtra::ggMarginal(p,type="density",margins="y",size=4)

#ggplot(plotdat[plotdat$Domain=="mus",],aes(variable,value,group=id))+geom_point(size=1,alpha=a)+geom_line(alpha=a,color=musGroupColor[plotdat$Domain=="mus"])+
#  ggtitle(paste0("Timepont changes in ",logStr,actVar))+ylab('Scores')+xlab('Time points')

```

Describe music change
```{r}
dmus=exp(dat$mus_T2)-exp(dat$mus_T1)
dmus_log = log(dmus)
dmus_log = dmus_log[is.finite(dmus_log)]

hist(dmus,100)
breaks=log(c(.05,1,3,10,30,100,300,1000,3000))
hist(dmus_log,breaks=breaks)

#Stops/Starts
hasMusT1=(dat$mus_T1 > 0)
hasMusT2=((dat$mus_T2 - dat$mus_T1) > 0)

nMusT1=sum(hasMusT1)
nT1=nrow(datbl)
nMusT2=sum(hasMusT2)
nT2=nrow(datfu)
nMusBoth=sum(hasMusT1&hasMusT2)

```
